---
title: "Foundations for statistical inference - Sampling distributions"
author: 'Emmanuel Kasigazi '
output:
  pdf_document: default
  html_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
set.seed(12124)
```

In this lab, you will investigate the ways in which the statistics from
a random sample of data can serve as point estimates for population
parameters. We're interested in formulating a *sampling distribution* of
our estimate in order to learn about the properties of the estimate,
such as its distribution.

::: {#boxedtext}
**Setting a seed:** We will take some random samples and build sampling
distributions in this lab, which means you should set a seed at the
start of your lab. If this concept is new to you, review the lab on
probability.
:::

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the
**tidyverse** suite of packages. We will also use the **infer** package
for resampling.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
library(tidyverse)
library(openintro)
library(tidyverse)
library(openintro)
library(dplyr)
library(tinytex)
```
```{r load-packagesaa, message=FALSE}
library(tidyverse)
library(openintro)
```
### The data

A 2019 Gallup report states the following:

> The premise that scientific progress benefits people has been embodied
> in discoveries throughout the ages -- from the development of
> vaccinations to the explosion of technology in the past few decades,
> resulting in billions of supercomputers now resting in the hands and
> pockets of people worldwide. Still, not everyone around the world
> feels science benefits them personally.
>
> **Source:** [World Science Day: Is Knowledge
> Power?](https://news.gallup.com/opinion/gallup/268121/world-science-day-knowledge-power.aspx)

The Wellcome Global Monitor finds that 20% of people globally do not
believe that the work scientists do benefits people like them. In this
lab, you will assume this 20% is a true population proportion and learn
about how sample proportions can vary from sample to sample by taking
smaller samples from the population. We will first create our population
assuming a population size of 100,000. This means 20,000 (20%) of the
population think the work scientists do does not benefit them personally
and the remaining 80,000 think it does.

```{r}
global_monitor <- tibble(
  scientist_work = c(rep("Benefits", 80000), rep("Doesn't benefit", 20000))
)
```

The name of the data frame is `global_monitor` and the name of the
variable that contains responses to the question *"Do you believe that
the work scientists do benefit people like you?"* is `scientist_work`.

We can quickly visualize the distribution of these responses using a bar
plot.

```{r bar-plot-pop1, fig.height=2.5, fig.width=10}
ggplot(global_monitor, aes(x = scientist_work)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "Do you believe that the work scientists do benefit people like you?"
  ) +
  coord_flip() 
```

We can also obtain summary statistics to confirm we constructed the data
frame correctly.

```{r summ-stat-pop, results = TRUE}
global_monitor %>%
  count(scientist_work) %>%
  mutate(p = n /sum(n))
```

## The unknown sampling distribution

In this lab, you have access to the entire population, but this is
rarely the case in real life. Gathering information on an entire
population is often extremely costly or impossible. Because of this, we
often take a sample of the population and use that to understand the
properties of the population.

If you are interested in estimating the proportion of people who don't
think the work scientists do benefits them, you can use the `sample_n`
command to survey the population.

```{r samp1}
samp1 <- global_monitor %>%
  sample_n(50)
```

This command collects a simple random sample of size 50 from the
`global_monitor` dataset, and assigns the result to `samp1`. This is
similar to randomly drawing names from a hat that contains the names of
all in the population. Working with these 50 names is considerably
simpler than working with all 100,000 people in the population.

1.  Describe the distribution of responses in this sample. How does it
    compare to the distribution of responses in the population.
    **Hint:** Although the `sample_n` function takes a random sample of
    observations (i.e. rows) from the dataset, you can still refer to
    the variables in the dataset with the same names. Code you presented
    earlier for visualizing and summarizing the population data will
    still be useful for the sample, however be careful to not label your
    proportion `p` since you're now calculating a sample statistic, not
    a population parameters. You can customize the label of the
    statistics to indicate that it comes from the sample.

```{r summ-stat-sample2, results = TRUE}
samp1 %>%
  count(scientist_work) %>%
  mutate(sample = n /sum(n))
```

scientist_work n sample <chr> <int> <dbl>

1 Benefits 40 0.8

2 Doesn't benefit 10 0.2

```{r bar-plot-pop2, fig.height=2.5, fig.width=10}
ggplot(samp1, aes(x = scientist_work)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "(Sample)Do you believe that the work scientists do benefit people like you?"
  ) +
  coord_flip() 
```

**In comparison it seems like its the same to me**

If you're interested in estimating the proportion of all people who do
not believe that the work scientists do benefits them, but you do not
have access to the population data, your best single guess is the sample
mean.

```{r phat-samp1}
samp1 %>%
  count(scientist_work) %>%
  mutate(p_hat = n /sum(n))
```

```{r inline-calc, include=FALSE}
# For use inline below
samp1_p_hat <- samp1 %>% 
  count(scientist_work) %>% 
  mutate(p_hat = n /sum(n)) %>% 
  filter(scientist_work == "Doesn't benefit") %>% 
  pull(p_hat) %>% 
  round(2)
```

Depending on which 50 people you selected, your estimate could be a bit
above or a bit below the true population proportion of `r samp1_p_hat`.
In general, though, the sample proportion turns out to be a pretty good
estimate of the true population proportion, and you were able to get it
by sampling less than 1% of the population.

2.  Would you expect the sample proportion to match the sample
    proportion of another student's sample? Why, or why not? If the
    answer is no, would you expect the proportions to be somewhat
    different or very different? Ask a student team to confirm your
    answer.

    **Yes, because samples usually tend to follow the population: There
    might be one or two oint difference but generally I expect the
    sample to follow / match mine**

**Insert your answer here**

3.  Take a second sample, also of size 50, and call it `samp2`. How does
    the sample proportion of `samp2` compare with that of `samp1`?
    Suppose we took two more samples, one of size 100 and one of
    size 1000. Which would you think would provide a more accurate
    estimate of the population proportion?

```{r samp2}
samp2 <- global_monitor %>%
  sample_n(50)
```

```{r summ-stat-sample1, results = TRUE}
samp2 %>%
  count(scientist_work) %>%
  mutate(sample2 = n /sum(n))
```

```         
  scientist_work      n sample2
```

<chr> <int> <dbl>

1 Benefits 39 0.78

2 Doesn't benefit 11 0.22

```{r bar-plot-pop3, fig.height=2.5, fig.width=10}
ggplot(samp2, aes(x = scientist_work)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "(Sample)Do you believe that the work scientists do benefit people like you?"
  ) +
  coord_flip() 
```

**The second sample is almost similar to the first, with just a few
points of difference, which don’t really make much of an impact. While
the first sample was 0.8, indicating some benefits, the second sample is
0.78, which is pretty close. A sample of 1,000 would provide a more
accurate depiction because it better represents and is closer to the
population.**

Not surprisingly, every time you take another random sample, you might
get a different sample proportion. It's useful to get a sense of just
how much variability you should expect when estimating the population
mean this way. The distribution of sample proportions, called the
*sampling distribution (of the proportion)*, can help you understand
this variability. In this lab, because you have access to the
population, you can build up the sampling distribution for the sample
proportion by repeating the above steps many times. Here, we use R to
take 15,000 different samples of size 50 from the population, calculate
the proportion of responses in each sample, filter for only the *Doesn't
benefit* responses, and store each result in a vector called
`sample_props50`. Note that we specify that `replace = TRUE` since
sampling distributions are constructed by sampling with replacement.

```{r iterate}
sample_props50 <- global_monitor %>%
                    rep_sample_n(size = 50, reps = 15000, replace = TRUE) %>%
                    count(scientist_work) %>%
                    mutate(p_hat = n /sum(n)) %>%
                    filter(scientist_work == "Doesn't benefit")
```

And we can visualize the distribution of these proportions with a
histogram.

```{r fig.show="hide"}
ggplot(data = sample_props50, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.02) +
  labs(
    x = "p_hat (Doesn't benefit)",
    title = "Sampling distribution of p_hat",
    subtitle = "Sample size = 50, Number of samples = 15000"
  )
```

Next, you will review how this set of code works.

4.  How many elements are there in `sample_props50`? Describe the
    sampling distribution, and be sure to specifically note its center.
    Make sure to include a plot of the distribution in your answer.

    **The dataset `sample_props50` contains 15,000 elements. Each
    element represents a sample proportion calculated from a sample of
    size 50 taken from the population, specifically for responses
    labeled "Doesn't benefit".**

    **Distribution characteristics:**

    **-Shape: The distribution appears approximately normal
    (bell-shaped) a normal distribution, which aligns with the Central
    Limit Theorem (CLT). Even though individual samples may vary, when
    we take many samples and compute their proportions, the distribution
    of those sample proportions approximates a normal curve.**

    **-Center: The peak of the distribution appears to be around 0.2 (or
    20%) suggesting that the mean of `p_hat` is close to 0.2.**

    **-Spread (Standard Error, SE): The variability in the histogram
    gives insight into the sampling variability. The standard deviation
    of this sampling distribution (Standard Error, SE) can be estimated
    using the formula:**

    SE=p(1−p)nSE = \sqrt{\frac{p(1 - p)}{n}}SE**=**np**(1−**p**)​​**

    **where** p≈0.2p \approx 0.2p**≈0.2 (from the histogram) and** n=50n
    = 50n**=50.**

**The sampling distribution demonstrates the variability we would expect
to see when taking repeated samples of size 50 from the population. The
fact that it's approximately normal is consistent with the Central Limit
Theorem, which states that the sampling distribution of a proportion
will be approximately normal for sufficiently large sample sizes.**

## Interlude: Sampling distributions

The idea behind the `rep_sample_n` function is *repetition*. Earlier,
you took a single sample of size `n` (50) from the population of all
people in the population. With this new function, you can repeat this
sampling procedure `rep` times in order to build a distribution of a
series of sample statistics, which is called the **sampling
distribution**.

Note that in practice one rarely gets to build true sampling
distributions, because one rarely has access to data from the entire
population.

Without the `rep_sample_n` function, this would be painful. We would
have to manually run the following code 15,000 times

```{r sample-code}
global_monitor %>%
  sample_n(size = 50, replace = TRUE) %>%
  count(scientist_work) %>%
  mutate(p_hat = n /sum(n)) %>%
  filter(scientist_work == "Doesn't benefit")
```

as well as store the resulting sample proportions each time in a
separate vector.

Note that for each of the 15,000 times we computed a proportion, we did
so from a **different** sample!

5.  To make sure you understand how sampling distributions are built,
    and exactly what the `rep_sample_n` function does, try modifying the
    code to create a sampling distribution of **25 sample proportions**
    from **samples of size 10**, and put them in a data frame named
    `sample_props_small`. Print the output. How many observations are
    there in this object called `sample_props_small`? What does each
    observation represent?

```{r Generate the sampling distribution with 25 samples of size 10}
sample_props_small <- global_monitor %>%
                    rep_sample_n(size = 10, reps = 25, replace = TRUE) %>%
                    count(scientist_work) %>%
                    mutate(p_hat = n /sum(n)) %>%
                    filter(scientist_work == "Doesn't benefit")
```

```{r Print the output}
sample_props_small
```

```{r visualizing  the smaller sampling distribution}
ggplot(data = sample_props_small, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.05) +
  labs(
    x = "p_hat (Doesn't benefit)",
    title = "Sampling distribution of p_hat",
    subtitle = "Sample size = 10, Number of samples = 25"
  )
```

**Looking at sample_props_small, there should be 25 observations in
total, because we specified reps = 25 in the code. Each observation in
sample_props_small represents one sample's results, specifically:**

**-It represents the proportion of "Doesn't benefit" responses (p_hat)
from a single random sample of 10 people from the population**

**-Each p_hat was calculated by taking the number of "Doesn't benefit"
responses in that sample of 10 people and dividing by 10**

**So each row in sample_props_small tells us: out of one random sample
of 10 people, what proportion said scientists' work "Doesn't benefit"
society. We did this sampling process 25 different times, which is why
we have 25 observations.**

## Sample size and the sampling distribution

Mechanics aside, let's return to the reason we used the `rep_sample_n`
function: to compute a sampling distribution, specifically, the sampling
distribution of the proportions from samples of 50 people.

```{r hist, fig.show='hide'}
ggplot(data = sample_props50, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.02)
```

The sampling distribution that you computed tells you much about
estimating the true proportion of people who think that the work
scientists do doesn't benefit them. Because the sample proportion is an
unbiased estimator, the sampling distribution is centered at the true
population proportion, and the spread of the distribution indicates how
much variability is incurred by sampling only 50 people at a time from
the population.

In the remainder of this section, you will work on getting a sense of
the effect that sample size has on your sampling distribution.

6.  Use the app below to create sampling distributions of proportions of
    *Doesn't benefit* from samples of size 10, 50, and 100. Use 5,000
    simulations. What does each observation in the sampling distribution
    represent? How does the mean, standard error, and shape of the
    sampling distribution change as the sample size increases? How (if
    at all) do these values change if you increase the number of
    simulations? (You do not need to include plots in your answer.)

**What each observation represents:**

Each observation in these sampling distributions represents one sample
proportion (p_hat) of people who responded "Doesn't benefit" from a
single random sample For sample size 10, each p_hat is calculated from
10 people For sample size 50, each p_hat is calculated from 50 people
For sample size 100, each p_hat is calculated from 100 people

Changes in the sampling distribution as sample size increases:

**Mean:**

The mean appears to stay relatively constant around 0.15-0.20 across all
sample sizes. This is expected since the mean of the sampling
distribution should equal the true population proportion

**Standard Error (spread of the distribution):**

The spread clearly decreases as sample size increases. Sample size 10:
Widest spread, with proportions varying greatly. Sample size 50:
Noticeably tighter spread. Sample size 100: Tightest spread This follows
the rule that standard error decreases with the square root of the
sample size

**Shape:**

Sample size 10: Discrete jumps, less normal-looking. Sample size 50:
More bell-shaped, approaching normal. Sample size 100: Most clearly
normal. This illustrates the Central Limit Theorem - as sample size
increases, the sampling distribution becomes more normal

**Effect of increasing number of simulations (currently 5000):**

**Mean:** Stays the same because the underlying population proportion
does not change. **Standard Error:** Does not change because SE depends
on **sample size**, not the number of simulations.\
**Shape:** The histogram becomes **smoother** as the number of
simulations increases because more observations provide a clearer
picture of the true distribution.\
More simulations would give a more precise picture of the sampling
distribution but wouldn't fundamentally change its characteristics. The
"jumpy" nature of the n=10 histogram would become smoother but would
still show discrete possible values

```{r shiny, echo=FALSE, eval=FALSE, results = TRUE}
# This R chunk will only run in interactive mode.
library(shiny)
shinyApp(
  ui <- fluidPage(
    
    # Sidebar with a slider input for number of bins 
    sidebarLayout(
      sidebarPanel(
        
        selectInput("outcome",
                    "Outcome of interest:",
                    choices = c("Benefits", "Doesn't benefit"),
                    selected = "Doesn't benefit"),
        
        numericInput("n_samp",
                     "Sample size:",
                     min = 1,
                     max = nrow(global_monitor),
                     value = 30),
        
        numericInput("n_rep",
                     "Number of samples:",
                     min = 1,
                     max = 30000,
                     value = 15000),
        
        hr(),
        
        sliderInput("binwidth",
                    "Binwidth:",
                    min = 0, max = 0.5,
                    value = 0.02,
                    step = 0.005)
        
      ),
      
      # Show a plot of the generated distribution
      mainPanel(
        plotOutput("sampling_plot"),
        textOutput("sampling_mean"),
        textOutput("sampling_se")
      )
    )
  ),
  
  server <- function(input, output) {
    
    # create sampling distribution
    sampling_dist <- reactive({
      global_monitor %>%
        rep_sample_n(size = input$n_samp, reps = input$n_rep, replace = TRUE) %>%
        count(scientist_work) %>%
        mutate(p_hat = n /sum(n)) %>%
        filter(scientist_work == input$outcome)
    })
    
    # plot sampling distribution
    output$sampling_plot <- renderPlot({
      
      ggplot(sampling_dist(), aes(x = p_hat)) +
        geom_histogram(binwidth = input$binwidth) +
        xlim(0, 1) +
        labs(
          x = paste0("p_hat (", input$outcome, ")"),
          title = "Sampling distribution of p_hat",
          subtitle = paste0("Sample size = ", input$n_samp, " Number of samples = ", input$n_rep)
        ) +
        theme(plot.title = element_text(face = "bold", size = 16))
    })
    
    ggplot(data = sample_props50, aes(x = p_hat)) +
      geom_histogram(binwidth = 0.02) +
      labs(
        x = "p_hat (Doesn't benefit)",
        title = "Sampling distribution of p_hat",
        subtitle = "Sample size = 50, Number of samples = 15000"
      )
    
    # mean of sampling distribution
    output$sampling_mean <- renderText({
      paste0("Mean of sampling distribution = ", round(mean(sampling_dist()$p_hat), 2))
    })
    
    # mean of sampling distribution
    output$sampling_se <- renderText({
      paste0("SE of sampling distribution = ", round(sd(sampling_dist()$p_hat), 2))
    })
  },
  
  options = list(height = 900) 
)
```

------------------------------------------------------------------------

## More Practice

So far, you have only focused on estimating the proportion of those you
think the work scientists doesn't benefit them. Now, you'll try to
estimate the proportion of those who think it does.

Note that while you might be able to answer some of these questions
using the app, you are expected to write the required code and produce
the necessary plots and summary statistics. You are welcome to use the
app for exploration.

7.  Take a sample of size 15 from the population and calculate the
    proportion of people in this sample who think the work scientists do
    enhances their lives. Using this sample, what is your best point
    estimate of the population proportion of people who think the work
    scientists do enchances their lives?

```{r benefits}
# Take a sample of size 15
samp_15 <- global_monitor %>%
  sample_n(15)

# Calculate proportion who think science benefits them
samp_15 %>%
  count(scientist_work) %>%
  mutate(p_hat = n /sum(n)) %>%
  filter(scientist_work == "Benefits")
```

```{r visualitzing}
# Visualize the sample distribution
ggplot(samp_15, aes(x = scientist_work)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "Do you believe that the work scientists do benefit people like you?",
    subtitle = "Sample size = 15"
  ) +
  coord_flip()
```

**The best point estimate for the population proportion of people who
think scientists' work "Benefits" them would be the sample proportion.
From the plot:**

**About 12 people responded "Benefits" out of the 15 people sampled
Therefore, p̂ = 12/15 = 0.80 or 80%**

**This sample proportion (0.80) is our best point estimate of the true
population proportion. Interestingly, this matches exactly with the true
population proportion that was used to create the original dataset
(80,000/100,000 = 0.80 or 80%), though in real-world sampling we
wouldn't typically know the true population value. While this is just
one sample and could vary from the true population proportion due to
sampling variability, it represents our best single guess at the
population parameter based on the available sample data**

8.  Since you have access to the population, simulate the sampling
    distribution of proportion of those who think the work scientists do
    enchances their lives for samples of size 15 by taking 2000 samples
    from the population of size 15 and computing 2000 sample
    proportions. Store these proportions in as `sample_props15`. Plot
    the data, then describe the shape of this sampling distribution.
    Based on this sampling distribution, what would you guess the true
    proportion of those who think the work scientists do enchances their
    lives to be? Finally, calculate and report the population
    proportion.

```{r distribution}
# Generate sampling distribution with 2000 samples of size 15
sample_props15 <- global_monitor %>%
                    rep_sample_n(size = 15, reps = 2000, replace = TRUE) %>%
                    count(scientist_work) %>%
                    mutate(p_hat = n /sum(n)) %>%
                    filter(scientist_work == "Benefits")

```

```{r visualtion}
# Create histogram of the sampling distribution
ggplot(data = sample_props15, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.05) +
  labs(
    x = "p_hat (Benefits)",
    title = "Sampling distribution of p_hat",
    subtitle = "Sample size = 15, Number of samples = 2000"
  )
```

**The distribution is roughly bell-shaped but appears slightly
left-skewed The peak is around 0.8-0.9 (80-90%) There's some natural
sampling variability, with proportions ranging from about 0.3 to 0.9
Most of the sample proportions are concentrated between 0.7 and 0.9**

**Best guess for true proportion: Based on this sampling distribution, I
would guess the true population proportion is around 0.8 (80%)
because:**

**This value is near the center of the distribution. The highest
concentration of sample proportions occurs around this value. The
sampling distribution should be centered at or near the true population
proportion**

**Actual population proportion**:

```{r}
global_monitor %>%
  count(scientist_work) %>%
  mutate(p = n /sum(n)) %>%
  filter(scientist_work == "Benefits")
```

**The true population proportion is 0.80 (80,000/100,000 = 0.8), which
aligns well with what the sampling distribution suggested. This
demonstrates how the sampling distribution of sample proportions can
give us good insight into the true population parameter, even with a
relatively small sample size of 15.**

9.  Change your sample size from 15 to 150, then compute the sampling
    distribution using the same method as above, and store these
    proportions in a new object called `sample_props150`. Describe the
    shape of this sampling distribution and compare it to the sampling
    distribution for a sample size of 15. Based on this sampling
    distribution, what would you guess to be the true proportion of
    those who think the work scientists do enchances their lives?

```{r sample 2K distr}
# Generate sampling distribution with 2000 samples of size 150
sample_props150 <- global_monitor %>%
                    rep_sample_n(size = 150, reps = 2000, replace = TRUE) %>%
                    count(scientist_work) %>%
                    mutate(p_hat = n /sum(n)) %>%
                    filter(scientist_work == "Benefits")

```

```{r}
# Create histogram of the sampling distribution
ggplot(data = sample_props150, aes(x = p_hat)) +
  geom_histogram(binwidth = 0.02) +
  labs(
    x = "p_hat (Benefits)",
    title = "Sampling distribution of p_hat",
    subtitle = "Sample size = 150, Number of samples = 2000"
  )
```

**Shape comparison:**

**Current distribution (n=150):**

**Much more symmetrical and clearly bell-shaped More concentrated around
the center (around 0.80) Narrower spread, ranging roughly from 0.70 to
0.90 Very close to a normal distribution**

**Previous distribution (n=15):**

**Was less symmetrical and showed some left skew Had more variability
and spread Ranged more widely from about 0.30 to 0.90 Was more irregular
in shape**

**Best guess for true proportion: Based on this sampling distribution
with n=150, I would estimate the true population proportion to be about
0.80 (80%) because:This is where the distribution centers and peaks .The
distribution is very symmetrical around this value With the larger
sample size of 150, we have more precision in our estimate than with
n=15. This increased precision is evident in the narrower spread of the
sampling distribution**

**This improved shape and precision illustrates why larger sample sizes
give us more reliable estimates of population parameters.**

10. Of the sampling distributions from 2 and 3, which has a smaller
    spread? If you're concerned with making estimates that are more
    often close to the true value, would you prefer a sampling
    distribution with a large or small spread?\

The sampling distribution with n=150 has a notably smaller spread than
the one with n=15. We can see this clearly by comparing the two
distributions:

n=150 distribution:Values mostly concentrated between 0.70 and 0.90.
Tight, compact shape. Clear peak around 0.80

n=15 distribution: Values spread widely from 0.30 to 0.90. Much more
dispersed shape. Less concentrated around the center

When making estimates, you would prefer a sampling distribution with a
smaller spread because: Sample proportions will be more consistently
close to the true population value. Less variability means more precise
estimates. Higher confidence that any single sample will give you a
value close to the true population proportion

This is why larger sample sizes (like n=150) are generally preferred -
they give us more precise estimates by reducing the variability in our
sampling distribution. With the smaller sample size (n=15), there's a
much higher chance of getting a sample proportion that's far from the
true population value.

------------------------------------------------------------------------
